PyPy's Approach to Virtual Machine Construction

### 2006-07-13 talk/dls2006/paper.tex
be more confident about, and give greater prominence to, the OOTyper

When the translation target is C or C-like environments, the first of
the transformation steps takes the RPython-typed flow graphs, still
containing polymorphic operations only, and produces flow graphs with
monomorphic C-like operations and C-like types.  In the simplest case,
this is the only transformation step: these graphs are directly fed to
the C emitting back-end, which turns them into ANSI C source code.

But RPython comes with automatic memory management, and this first
transformation step produces flow graphs that also assume automatic
memory management.  Generating C code directly from there produces a
fully leaking program, unless we link it with an external garbage
collector (GC) like the Boehm conservative GC \cite{Boehm}\cite{boehm-softw},
which is a viable option.

We have two alternatives, each implemented as a transformation step.
The first one inserts naive reference counting throughout the whole
program's graphs, which without further optimizations gives a somewhat
bad performance (it should be noted that the CPython interpreter is also
based on reference counting, and experience suggests that it was not a
bad choice in this particular case).

The other, and better, alternative is an exact GC, coupled with a
transformation, the \textit{GC transformer}.  It inputs C-level-typed graphs
and replaces all \texttt{malloc} operations with calls to a garbage
collector's allocation routine.
The transformation inspects all the graphs to discover the
\texttt{struct} types in use by the program, and assigns a unique type id to
each of them.  These type ids are then collected in internal tables that
describe the layout of the structures, e.g.\ their sizes and the location
of the pointer fields.

@@

%%% XXX can we be more optimistic about the OOTyper?  It's pretty
%%% likely that pypy.net will work in some form by OOPSLA -- hi
%%% Antonio :)

Finally, currently under development is a variant of the very first
transformation step, for use when targeting higher-level,
object-oriented (OO) environments.  It is currently being designed
together with back-ends for Smalltalk/Squeak\footnote{Our simple OO
type system is designed for \textit{statically-typed} OO environments,
including Java; the presence of Smalltalk as a back-end might be
misleading in that respect.} and CLI/.NET.  This first transformation
step, for C-like environments, is called the \textit{LLTyper} or
low-level typer: it produces
C-level flow graphs, where the object-oriented features of RPython
(classes and instances) become manipulations of C structs with
explicit virtual table pointers.  By contrast, for OO environments the
transformation step is called the \textit{OOTyper:} it targets a simple
object-oriented type system, and preserves the classes and instances
of the original RPython program.  The LLTyper and OOTyper still have
much code in common, to convert the more Python-specific features like
its complex calling conventions.

@@

Current work also includes the OOTyper and back-ends for CLI/.NET and
Smalltalk/Squeak, all in-progress.

>>>>

The first of the transformation steps takes the RPython-typed flow
graphs, still containing polymorphic operations only, and produces
flow graphs with operations more familiar to the target environment.
When the target is C or a C-like environment, this means monomorphic
C-like operations and C-like types.  In the simplest case, this is the
only transformation step: these graphs are directly fed to the C
emitting back-end, which turns them into ANSI C source code.

Not all targets of interest are C-like, though, and we have recently
developed a variant of this step for use when targeting higher-level,
object-oriented (OO) environments.  Its design was motivated by work
on back-ends for Smalltalk/Squeak\footnote{Our simple OO type system
  is designed for \textit{statically-typed} OO environments, including
  Java; the presence of Smalltalk as a back-end might be misleading in
  that respect.} and CLI/.NET.  When targeting a C-like environment,
the first transformation step is called the \textit{LLTyper} or
low-level typer: it produces C-level flow graphs, where the
object-oriented features of RPython (classes and instances) become
manipulations of C structs with explicit virtual table pointers.  By
contrast, for OO environments this step is called the
\textit{OOTyper:} it targets a simple object-oriented type system, and
preserves the classes and instances of the original RPython program.
The LLTyper and OOTyper still have much code in common, to convert the
more Python-specific features like the more complex calling
conventions.  In the sequel, we assume the use of the LLTyper, as
there are as yet no back-ends for the OOTyper which are complete
enough to translate the Standard Interpreter -- although the CLI
backend should be able to within a few months.

For an example of a transformation that is applied after this first
step, consider that RPython has automatic memory management. Even with
the LLTyper, the first transformation step produces flow graphs that
also assume automatic memory management.  Generating C code directly
from there produces a fully leaking program, unless we link it with an
external garbage collector (GC) like the Boehm conservative GC
\cite{Boehm}\cite{boehm-softw}, which is a viable option.

We have two other alternatives, each implemented as a transformation step.
The first one inserts naive reference counting throughout the whole
program's graphs, which without further optimizations gives a somewhat
bad performance (it should be noted that the CPython interpreter is also
based on reference counting, and experience suggests that it was not a
bad choice in this particular case).

The other, and better, alternative is an exact GC, coupled with a
transformation, the \textit{GC transformer}.  It takes as input
C-level-typed graphs and replaces all \texttt{malloc} operations with
calls to a garbage collector's allocation routine.  The transformation
inspects all the graphs to discover the structure types in use
by the program, and assigns a unique type id to each of them.  These
type ids are then collected in internal tables that describe the
layout of the structures, e.g.\ their sizes and the location of the
pointer fields.

@@

@@

Current work also includes refininments of the OOTyper and work on
back-ends that use it, including those targeting CLI/.NET and
Smalltalk/Squeak.
